Split-Apply-Combine with Collapsing Groups

### 2022-12-29 paper/article.Rnw
writing in progress

In all packages mentioned, the calculation for each group uses data available
within the group. However, there are valid use cases where a group aggregate is
determined using attributes from out-of-group entities. One example where this
occurs is in the area of small area estimation (see e.g.
\citet{molina2015sae}. Here, one wishes to estimate an aggregate for a group,
for example a geographical region, or a detailed population subset, where the
number of (sampled) observations so small that the variance of the estimate
would be unacceptably large. Small area estimation is a collection of
bias-variance trade-offs where one `borrows statistical strength' from
out-of-group records to obtain a stable estimate for groups with few
observations. The out-of-group records can be obtained, for example by
combining the original group with a group of records that are deemed similar in
certain respects. A second area where out-of-group records play a role is in
certain hot-deck imputation methods \cite{andridge2010review}.  In
$k$-nearest-neighbours imputation for example, one finds a set of $k$ donor
records that are preferably in the same group, but this condition may be
relaxed if there are not enough records in the group. In the \pkg{VIM} package
for \proglang{R} \citep{kowarik2016imputation}, this is controlled by a
combination of the Gower distance and setting conditions on minimal number of
donors. A third example comes from the field of statistical disclosure control
\citep{hundepool2012statistical}. In this field, the so-called $p$\% rule states
that the value in a grouped aggregation may not be determined for more than $p$
percent by a single entity. This occurs for example, when a certain economic
activity in a geographic region is dominated by a single company. In such cases
the value can either not be published, or one publishes an aggregate over
larger groups.

>>>>

In all packages mentioned, the calculation for each group uses data available
within the group. However, there are valid use cases where a group aggregate is
determined using attributes from out-of-group entities. One example where this
occurs is in the area of small area estimation (see e.g.
\citet{molina2015sae}. Here, one wishes to estimate an aggregate for a group,
for example a geographical region, or a detailed population subset, where the
number of (sampled) observations so small that the variance of the estimate
would be unacceptably large. Small area estimation is a collection of
bias-variance trade-offs where one `borrows statistical strength' from
out-of-group records to obtain a stable estimate for groups with few
observations. The out-of-group records can be obtained, for example by
combining the original group with a group of records that are deemed similar in
certain respects. A second area where out-of-group records play a role is in
certain hot-deck imputation methods \citep{andridge2010review}.  In
$k$-nearest-neighbours imputation for example, one finds a set of $k$ donor
records that are preferably in the same group, but this condition may be
relaxed if there are not enough records in the group.  In the \pkg{VIM} package
for \proglang{R} \citep{kowarik2016imputation}, this is controlled by a
combination of the Gower distance and setting conditions on minimal number of
donors. In practice, imputation is often performed via a fall-though scenario,
where one first tries to estimate a model within a group, but if the group is
too small for a reliable estimate of model parameters, the group is enlarged by
combining similar groups, similar to the small-area estimation scenario. A
third example comes from the field of statistical disclosure control
\citep{hundepool2012statistical}. In this field, the so-called $p$\% rule
states that the value in a grouped aggregation may not be determined for more
than $p$ percent by a single entity. This occurs for example, when a certain
economic activity in a geographic region is dominated by a single company. In
such cases the value can either not be published (suppression), or one
publishes an aggregate over larger groups.
